---
title: 'Lab6: Clustering'
author: "Nick McManus"
date: "2023-02-16"
output: 
  html_document:
    code_folding: show
---

```{r setup, echo = TRUE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(tidyverse)
library(janitor)
library(palmerpenguins)

## packages for cluster analysis
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```



# Intro to cluster analysis - k-means, hierarchical

## Part 1: K-means clustering

```{r}
ggplot(penguins) +
  geom_point(aes(x=bill_length_mm, y=bill_depth_mm, color = species, shape = sex),
             size = 3, alpha = 0.7) +
  scale_color_manual(values = c('orange','cyan4','darkmagenta')) +
  theme_minimal()


ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = species,
                 shape = sex),
             size = 3, alpha = 0.7) +
  scale_color_manual(values = c('orange','cyan4','darkmagenta')) +
  theme_light()
```


### Create a complete, scaled version of the data

```{r}
penguins_complete <- penguins %>% 
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm)

penguins_scale <- penguins_complete %>% 
  select(ends_with('_mm'), body_mass_g) %>% 
  scale()
```


### Estimate number of clusters

```{r}
number_est <- NbClust(penguins_scale, min.nc = 2, max.nc = 10, method = 'kmeans')

fviz_nbclust(penguins_scale, FUNcluster = kmeans, method = 'wss', k.max = 10)
```



### run some k-means

```{r}
set.seed(123)
penguins_km <- kmeans(penguins_scale,
                      centers = 3,
                      iter.max = 10,  #default value
                      nstart = 25)

#penguins_km$size
#penguins_km$cluster
penguins_cl <- penguins_complete %>% 
  mutate(cluster_no = factor(penguins_km$cluster))
```


```{r}
ggplot(penguins_cl) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = cluster_no,
                 shape = species))

ggplot(penguins_cl) +
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm,
                 color = cluster_no,
                 shape = species)) +
  scale_color_viridis_d()  #d for discrete


penguins_cl %>% 
  select(species, cluster_no) %>% 
  table()
```


## Hierarchical clustering

### Start with complete linkage

```{r}
# create distance matrix
peng_dist <- dist(penguins_scale, method = 'euclidean')

# hierarchical clustering (complete linkage)
peng_hc_complete <- hclust(peng_dist, method = 'complete')
##other methods: single, average, ward.D

# plot a dendrogram
plot(peng_hc_complete, cex = .6, hang = -1)

# cut tree into three clusters
peng_cut_hc <- cutree(peng_hc_complete, 3)
table(peng_cut_hc, penguins_complete$species)
```



## World Bank data: read in and simplify

```{r}
wb_env <- read_csv(here::here('data/wb_env.csv'))

wb_ghg_20 <- wb_env %>% 
  slice_max(ghg, n = 20)

summary(wb_ghg_20)

wb_scaled <- wb_ghg_20 %>% 
  select(3:7) %>%  #could also do variable names
  scale()

summary(wb_scaled)

# turned to matrix so lost rownames, can add from og dataset
rownames(wb_scaled) <- wb_ghg_20$name
```


### Find Euclidean distances 

```{r}
euc_distance <- dist(wb_scaled, method = 'euclidean')
```

### Perform hierarchical clustering w/complete linkage

```{r}
hc_complete <- hclust(euc_distance, method = 'complete')
plot(hc_complete, cex = 0.6, hang = -1)
```


### Perform hierachical clustering by single linkage

```{r}
hc_single <- hclust(euc_distance, method= 'single')
plot(hc_single, cex = 0.9, hang = -1)
```


### Make a tanglegram

Compare two dendrograms at once
```{r}
dend_complete <- as.dendrogram(hc_complete)
dend_single <- as.dendrogram(hc_single)

tanglegram(dend_complete, dend_single)

entanglement(dend_complete, dend_single)

untangle(dend_complete, dend_single, method = 'step1side') %>% 
  entanglement()

untangle(dend_complete, dend_single, method = 'step1side') %>% 
  tanglegram(common_subtrees_color_branches = TRUE)
```


### Dendrogram in ggplot
```{r}
ggdendrogram(hc_complete, rotate = TRUE)+
  theme_minimal() +
  labs(x = 'Country')
```














